Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-02-28 report/paper.tex
fix snafus added in merge

{
An AI assistant is a component
of the system that 
guides data analysts through a single data collection, integration, preparation, 
analytics, or reporting task.
For example, an assistant might
specialize in identifying suspicious
values, such as "-1" in a column labelled
"heart rate", which might indicate
that certain data was not collected.
Assistants are interactive, in that
they propose a transformation
of the data to the analyst, based on combinations
of statistical and symbolic AI analysis,
and then refine the transformation
based on feedback from the analyst.
Typically analysts will interact with many separate assistants over 
the course of an analysis, each of which specialize in a single kind of task.
}
\subsection{Architecture of AI assistants}

There are two important design aspects of AI assistants, which may seem 
contradictory.
 First, they must be interactive, following
 the first principle behind Wrattler, but
 they must also be reproducible.
We square this circle by using the notebook
system as a reproducible interlingua. 
The output of an AI assistant is always code, rather
than being a black-box that transforms data,
and it is this code that is agreed between the AI and the analyst during the interaction.
Thus the interaction is viewed as a ``design phase''
that is not reproduced on new data.

@@

\paragraph{Domain specific languages.}
The output of an assistant is code in a small domain-specific language designed for the task that
the AI assistant is solving. The code can be used in two ways -- during the interaction with the
AI assistant, the code is exposed through a Wrattler DSL framework as discussed in 
Section~\ref{sec:wrattler-ai}. For production use, the code is translated to a language like R or 
Python. A typical code produced by an AI assistant will implement a data transformation or
produce a visualization.

\paragraph{Interactivity.}
Assistants should usually contain user interaction in their design, and adapt rapidly to feedback 
from the analyst. For example, a data cleaning assistant would probably want to ask the user to 
confirm potential changes, and if the user says no, to reconfigure the cleaning module 
significantly so that the user does not need to keep rejecting similar changes.

However, at the same time, our data preparation tools should be usable in "batch mode" by analysts 
who already know what they want to do and don't want to mess with a wizard. We should aim to 
produce the equivalent of scikit-learn for data preparation.

The output snippet should be readable, and something that a data analyst can look at to verify 
that the transformation accomplishes what they had wanted. As far as possible, the snippet 
should be "reusable" in a sense that it could be copy-pasted into another script that was 
processing a similar dataset and would still work. You can think of a snippet as something 
that a person might write in a single cell of a Jupyter notebook.

\vspace{1em}
AI assistants have access to information in the Wrattler data store and can use two separate
aspects for their guidance:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.
\vspace{0.25em}

\item[--] AI assistants can access the dependency graph created for a 
notebook. This allows the them to access the transformation history, which is the code 
that has already been run during the analysis. The idea is that assistants may need to know the 
context of transformations that have already been performed in order to decide what to suggest next.
\end{itemize}

The assistants should be self-contained in the sense that they interact with each other
only through the 
data store. This restriction is necessary in order
for provenance tracking and reproducibility.

@@

\vspace*{1em}

@@

{\color{red}
\textbf{TP: My rough notes from discussion with JG}

~

\noindent
Data store stores data frames with metadata. This is a good start for data science work,
because most languages can provide mappings to/from data frames.
}

\subsection{Semantics annotations}
{\color{red}
Data store provides a way of annotating cells, columns, combinations of columns and rows.
For example, a column may be annotated with a type that indicates that its values are a mix of
postcodes and city names. A cell can be annotated with an information saying that this value is 
more likely a postcode than a city name. We also need to support annotaitons on a combination of
columns, because three columns can jointly represent an address. A row can be annotated with a 
provenance -- for example, when merging rows from two data sets with different sources.
}

>>>>

{
An AI assistant is a component
of the system that
guides data analysts through a single data collection, integration, preparation,
analytics, or reporting task.
For example, an assistant might
specialize in identifying suspicious
values, such as "-1" in a column labelled
"heart rate", which might indicate
that certain data was not collected.
Assistants are interactive, in that
they propose a transformation
of the data to the analyst, based on combinations
of statistical and symbolic AI analysis,
and then refine the transformation
based on feedback from the analyst.
Typically analysts will interact with many separate assistants over
the course of an analysis, each of which specialize in a single kind of task.

\subsection{Architecture of AI assistants}

There are two important design aspects of AI assistants, which may seem
contradictory.
 First, they must be interactive, following
 the first principle behind Wrattler, but
 following the second,
 they must also be reproducible.
We square this circle by using the notebook
system as a reproducible interlingua.
The output of an AI assistant is always code, rather
than being a black-box that transforms data,
and it is this code that is agreed between the AI and the analyst during the interaction.
Thus the interaction is viewed as a ``design phase''
that is not reproduced on new data.

More specifically, an AI assistant defines a domain specific language (DSL) for the task at hand. When invoked from
a notebook, it guides the user in creating a script in the DSL. Running the script performs
the required operation (such as data cleaning). The assistant defines the semantics
of those operations and, when the script is executed, stores the result into a data
store. This new data frame can contain new data or additional information (such as their
inferred probabilistic type).
The interaction is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
store and returns suggestions to the notebook, which the user can review and accept.

@@

\paragraph{Domain specific languages.}

Wrattler includes a scripting language with simple syntax that is used by individual AI assistants
to define domain specific languages. The syntax is shared across multiple AI assistnts, but the
operations are provided by individual assistants.

For example, an AI assistant based on datadiff
guides the user in writing scripts that transform a malformed data frame into a correct format
specified by a correct data frame. The resulting script might look as follows:
%
\begin{equation*}
\begin{array}{l}
\ident{datadiff(broadband2014, broadband2015)}\\
\quad.\ident{drop\_column}(\str{WT\_national})\\
\quad.\ident{drop\_column}(\str{WT\_ISP})\\
\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
\end{array}
\end{equation*}
%
The script specifies that two columns should be dropped from the badly structured data frame and
one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
\strf{Urban} and \strf{Rural}).

The code produced by an AI assistant can be used in two ways. First, the analyst writes the code
interactively during an interaction with the AI assistant. During this phase, Wrattler
keeps a fine-grained representation of the code in the dependency graph (second cell in
Figure~\ref{fig:deps}), which allows Wrattler to provie live previews of the result, giving the
user an immediate feedback.

At a later stage, when the code is used in production, the code can be translated to a language
like R or Python. This translation uses the fine-grained representation of the code in the
dependency graph and maps individual nodes to corresponding Python or R functions.

\paragraph{Interactivity.}

Assistants should usually contain user interaction in their design, and adapt rapidly to feedback
from the analyst. For example, a data cleaning assistant would probably want to ask the user to
confirm potential changes, and if the user says no, to reconfigure the cleaning module
significantly so that the user does not need to keep rejecting similar changes.

In the above example, the interaction is provided through the ``.'' operator of the scripting
language. When the analyst types the first line to invoke the AI assistant and types ``.''
the datadiff assistant offers the most likely patches and the user can choose which ones
to accept. Scripts constructed in such way are readable, allowing the data analyst to look at them
and verify that the transformations accomplish the intended task (such verification further
benefits from live previews that Wrattler provides when the analyst navigates through the
individual transformations).

At the same time, AI assistants can be used in a fully automatic ``batch mode'' by analysts
who already know what they want to achieve and know that the assistant produces the result they
require. In this case, Wrattler will simply automatically accept the default recommendations made
by an AI assistant. This allows packaging the assistant as an equivalent of scikit-learn for
data preparation.

\vspace{1em}
The Wrattler architecture supports the development of AI assistants in a number of ways. First, AI
assistants have access to the data store and can use it to provide guidance about both data and
code. Second, Wrattler includes an extensible framework for defining domain-specific languages
(DSLs) that can describe, for example, data extraction, cleaning, transformation or visualization.
The data store provides AI assistants with
two sources of information that they can exploit
to determine what guidance to provide the analyst:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.

\item[--] AI assistants can access the dependency graph created for a
notebook. This allows the them to access the transformation history, which is the code
that has already been run during the analysis. The gives assistants information about the
context such as transformations that have already been performed.
\end{itemize}

\medskip \noindent
The assistants should be self-contained in the sense that they interact with each other
only through the
data store. This restriction is necessary in order
for provenance tracking and reproducibility.

@@

@@

The data store links together individual components of Wrattler. It stores the raw input
data for data analysis such as downloaded web pages, raw data frames and also data frames
produced as the result of running a script or invoking an AI assistant. In addition to data,
the data store also stores multiple versions of the dependency graph created by Wrattler.

In the initial version of Wrattler, we restrict the data storage to raw input files and
data frames, however we note that it would be desirable to include other data structures
such graph or network data.

In this section, we focus on the initial functionality that is required from the data store.
We focus on the public interface exposed by the data store, but We do not discuss the technolgy
used for data storage -- this can be decided during the development.

Data frames stored in the data store can be annotated with metadata. Such metadata can
communicate information between multiple AI assistants. For example, first AI assistant can
infer that a type of a pair of columns represents GPS coordinate and produce a data frame
with annotations specifying this information. A second AI assistant can then use this information
and automatically provide a data visualization in the form of map. In the rest of this section,
we discuss the structure of those annotations.

\subsection{Semantic annotations}
As noted in Section~\ref{sec:wrattler-ds}, annotations can be attached to individual cells,
entire columns and rows, but also combinations of multiple columns or rows. As an example,
consider the following small table that records data on broadband speed (and illustrates a number
of real-world problems):

\vspace{1em}
\noindent
\begin{tabular}{llc}
\toprule
\textbf{ISP}\qquad\qquad\qquad & \textbf{Postcode}\qquad\qquad & \textbf{Download} \\
\midrule
Virgin & CB43BE & 46.34 \\
BT & CB43PT & N/A \\
Virgin & NW53HL & 41.2 \\
BT & Tottenham & 16.4 \\
\bottomrule
\end{tabular}
\vspace{1em}

\noindent
If we run an AI asistant that performs probabilistic type inference (ptype) on this data set,
it will annotate the table with a number of information:
\begin{itemize}
\item[--] aa
\end{itemize}

{\color{red}
Data store provides a way of annotating cells, columns, combinations of columns and rows.
For example, a column may be annotated with a type that indicates that its values are a mix of
postcodes and city names. A cell can be annotated with an information saying that this value is
more likely a postcode than a city name. We also need to support annotaitons on a combination of
columns, because three columns can jointly represent an address. A row can be annotated with a
provenance -- for example, when merging rows from two data sets with different sources.
}
